---
title: "US Presidential Election Results (1976 - 2020)"
format:
  html:
    code-fold: true
    code-tools: true
runtime: shiny
execute:
  echo: true
editor: 
  markdown: 
    wrap: 72
---
## Dashboard {.tabset .tabset-pills .nav-stacked}

```{r setup, include=FALSE}
library(shiny)
library(tidyverse)
library(plotly)
library(pander)
```

### General Interactive Map

```{r, echo = FALSE}
#| panel: sidebar
sliderInput("year", "Select Election Year:",
            min = 1976, max = 2020,
            value = 2020, step = 4, sep = "")
```

```{r, echo = FALSE, warning = FALSE, message = FALSE}
#| context: server
library(shiny)
library(tidyverse)
library(plotly)
library(maps)

# Load US map and data
us_map <- map_data("state")
decadesdata <- read_csv("C:/Users/micha/OneDrive/Data Wrangling & Visualization/Week 9/Task Data Search/1976-2020-president.csv")

election_winners <- tribble(
  ~year, ~winner, ~matchup,
  1976, "CARTER, JIMMY (DEMOCRAT)", "CARTER, JIMMY vs. FORD, GERALD",
  1980, "REAGAN, RONALD (REPUBLICAN)", "REAGAN, RONALD vs. CARTER, JIMMY",
  1984, "REAGAN, RONALD (REPUBLICAN)", "REAGAN, RONALD vs. MONDALE, WALTER",
  1988, "BUSH, GEORGE H. W. (REPUBLICAN)", "BUSH, GEORGE H. W. vs. DUKAKIS, MICHAEL",
  1992, "CLINTON, BILL (DEMOCRAT)", "CLINTON, BILL vs. BUSH, GEORGE H. W.",
  1996, "CLINTON, BILL (DEMOCRAT)", "CLINTON, BILL vs. DOLE, BOB",
  2000, "BUSH, GEORGE W. (REPUBLICAN)", "BUSH, GEORGE W. vs. GORE, AL",
  2004, "BUSH, GEORGE W. (REPUBLICAN)", "BUSH, GEORGE W. vs. KERRY, JOHN",
  2008, "OBAMA, BARACK (DEMOCRAT)", "OBAMA, BARACK vs. MCCAIN, JOHN",
  2012, "OBAMA, BARACK (DEMOCRAT)", "OBAMA, BARACK vs. ROMNEY, MITT",
  2016, "TRUMP, DONALD J. (REPUBLICAN)", "CLINTON, HILLARY vs. TRUMP, DONALD J.",
  2020, "BIDEN, JOSEPH R. (DEMOCRAT)", "BIDEN, JOSEPH R. vs. TRUMP, DONALD J."
)


# Define color scale
rating_colors_shaded <- c(
  "Safe DEMOCRAT" = "#244999",
  "Likely DEMOCRAT" = "#577CCC",
  "Lean DEMOCRAT" = "#8AAFFF",
  "Tilt DEMOCRAT" = "#949BB3",
  "Safe REPUBLICAN" = "#D22532",
  "Likely REPUBLICAN" = "#FF5865",
  "Lean REPUBLICAN" = "#FF8B98",
  "Tilt REPUBLICAN" = "#CF8980",
  "Missing" = "#D3D3D3"
)

# Function to process election data
process_year <- function(yr, data) {
  results <- data %>%
    filter(year == yr, party_simplified %in% c("DEMOCRAT", "REPUBLICAN")) %>%
    group_by(state, party_simplified) %>%
    summarise(votes = sum(candidatevotes), .groups = "drop") %>%
    pivot_wider(names_from = party_simplified, values_from = votes, values_fill = 0) %>%
    mutate(
      total_votes = DEMOCRAT + REPUBLICAN,
      margin = abs(DEMOCRAT - REPUBLICAN) / total_votes,
      winner = if_else(REPUBLICAN > DEMOCRAT, "REPUBLICAN", "DEMOCRAT"),
      margin_label = case_when(
        margin >= 0.15 ~ "Safe",
        margin >= 0.10 ~ "Likely",
        margin >= 0.05 ~ "Lean",
        TRUE ~ "Tilt"
      ),
      rating_category = paste(margin_label, winner),
      rating_category = ifelse(rating_category %in% names(rating_colors_shaded), rating_category, "Missing"),
      fill_color = as.character(rating_colors_shaded[rating_category]),
      tooltip = paste0(
        toupper(state), "<br>",
        "Democrat: ", formatC(DEMOCRAT, format = "d", big.mark = ","), "<br>",
        "Republican: ", formatC(REPUBLICAN, format = "d", big.mark = ","), "<br>",
        "Winner: ", winner, "<br>",
        "Margin: ", scales::percent(margin)
      ),
      region = tolower(state)
    )
  
  df <- left_join(us_map, results, by = "region") %>%
    filter(!is.na(tooltip)) %>%  # 🚨 This removes stray polygon ghosts
    group_by(region) %>%
    tidyr::fill(tooltip, fill_color, .direction = "downup") %>%
    ungroup()
  df <- df %>%
    mutate(
      tooltip = ifelse(is.na(tooltip), paste0(toupper(region), "<br>No data"), tooltip),
      fill_color = ifelse(is.na(fill_color), "#D3D3D3", fill_color)
    )
  return(df)
}
```

```{r election_map, echo = FALSE}
# Plotly render
renderPlotly({
  year_selected <- as.numeric(input$year)
  df <- process_year(year_selected, decadesdata)

   # 1. State-level winners
 state_winners <- decadesdata %>%
  filter(year == year_selected, party_simplified %in% c("DEMOCRAT", "REPUBLICAN")) %>%
  group_by(state, candidate, party_simplified) %>%
  summarise(votes = sum(candidatevotes), .groups = "drop") %>%
  group_by(state) %>%
  slice_max(order_by = votes, n = 1, with_ties = FALSE) %>%
  distinct(state, .keep_all = TRUE) %>%  
  mutate(year = year_selected)


title_data <- election_winners %>% filter(year == year_selected)

title_text <- paste0(
  "Election Results - ", year_selected
)

# Plot
plot_ly(
  data = df,
  type = "scatter",
  mode = "none",
  x = ~long,
  y = ~lat,
  split = ~group,
  fill = "toself",
  fillcolor = ~fill_color,
  text = ~tooltip,
  hoveron = "fills",
  hoverinfo = "text",
  hoverlabel = list(
    font = list(size = 20, color = "white")
  ),
  line = list(color = "black", width = 0.5),
  showlegend = FALSE
) %>%
  layout(
    title = list(
      text = title_text,
      x = 0.5,
      xanchor = "center",
      font = list(size = 15)
    ),
    xaxis = list(visible = FALSE),
    yaxis = list(visible = FALSE, scaleanchor = "x")
  )
})
```

This interactive map visualizes U.S. presidential election outcomes from
1976 to 2020. Use the slider to select a year and view which party won
each state. Color intensity indicates the margin of victory: darker
shades represent decisive wins, while lighter shades highlight more
competitive races. Hover over states for vote counts, margins, and party
breakdowns.

*States are shaded by party and competitiveness — darker = larger
margin, lighter = tighter race. Use the slider to explore results year
by year.*

### Bellwether States Analysis

<br> First, we are going to have to determine who won the Presidential
Election every year from 1976 to 2020. This is only 12 elections, so
there are minimal amounts of joins. <br>

```{r, echo = FALSE}
election_winners <- tribble(
  ~year, ~winner_party,
  1976, "DEMOCRAT",
  1980, "REPUBLICAN",
  1984, "REPUBLICAN",
  1988, "REPUBLICAN",
  1992, "DEMOCRAT",
  1996, "DEMOCRAT",
  2000, "REPUBLICAN",
  2004, "REPUBLICAN",
  2008, "DEMOCRAT",
  2012, "DEMOCRAT",
  2016, "REPUBLICAN",
  2020, "DEMOCRAT"
)

state_results <- decadesdata %>%
  filter(party_simplified %in% c("DEMOCRAT", "REPUBLICAN")) %>%
  group_by(year, state, party_simplified) %>%
  summarise(votes = sum(candidatevotes), .groups = "drop") %>%
  group_by(year, state) %>%
  slice_max(order_by = votes, n = 1, with_ties = FALSE) %>%
  ungroup()

bellwether_analysis <- state_results %>%
  left_join(election_winners, by = "year") %>%
  mutate(
    match_national = if_else(party_simplified == winner_party, 1, 0)
  )
DT::datatable(bellwether_analysis)


bellwether_summary <- bellwether_analysis %>%
  group_by(state) %>%
  summarise(
    times_matched = sum(match_national),
    total_elections = n(),
    match_rate = round(100 * times_matched / total_elections, 1)
  ) %>%
  arrange(desc(times_matched))

bellwether_summary %>%
  top_n(10, times_matched) %>%
  ggplot(aes(x = reorder(state, times_matched), y = times_matched, fill = state)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  labs(
    title = "Top 15 Bellwether States (1976–2020)",
    x = "State",
    y = "Times Matched with National Winner"
  ) +
  theme_minimal() +
  theme(legend.position = "none")  

```

<br> Since 1976, Ohio Pennslyvania, Nevada, and Florida have aligned
with the winners of the general election the most. Interestingly enough,
states like Tennessee, New Mexico, Missouri, Louisiana, Kentucky, and
Delaware also have matched with the national winner, even though these
states are not considered swing states today. If we are to use data
since 1976, seeing how these states vote in a given election should give
us a good idea as to who won the election.

### Most Competitive States

To determine which states are the most competitive, we are going to have
to generate a column that subtracts the differences between the matchups
in each state within each year and extracts a margin in both raw votes
and as a percentage.

<br> **Most Competitive by raw margins** <br>

```{r, warning = FALSE, echo = FALSE}
state_margins <- decadesdata %>%
  filter(party_simplified %in% c("DEMOCRAT", "REPUBLICAN")) %>%
  group_by(year, state, party_simplified) %>%
  summarise(votes = sum(candidatevotes), .groups = "drop") %>%
  pivot_wider(names_from = party_simplified, values_from = votes, values_fill = 0) %>%
  mutate(
    total_votes = DEMOCRAT + REPUBLICAN,
    abs_margin = abs(DEMOCRAT - REPUBLICAN),
    pct_margin = abs_margin / total_votes * 100,
    winner = if_else(REPUBLICAN > DEMOCRAT, "REPUBLICAN", "DEMOCRAT")
  ) %>%
  select(year, state, winner, abs_margin, pct_margin)
top10_abs <- state_margins %>%
  group_by(state) %>%
  summarise(mean_abs_margin = mean(abs_margin, na.rm = TRUE)) %>%
  slice_min(mean_abs_margin, n = 10) %>%
  pull(state)

top10_abs_summary <- state_margins %>%
  group_by(state) %>%
  summarise(mean_abs_margin = mean(abs_margin, na.rm = TRUE)) %>%
  slice_min(mean_abs_margin, n = 10)

top10_median_abs <- state_margins %>%
  group_by(state) %>%
  summarise(median_abs_margin = median(abs_margin, na.rm = TRUE)) %>%
  arrange(median_abs_margin) %>%
  slice(1:10)

state_margins %>%
  filter(state %in% top10_median_abs$state) %>%
  mutate(state = factor(state, levels = top10_median_abs$state)) %>%
  ggplot(aes(x = state, y = abs_margin, fill = state)) +
  geom_boxplot() +
  coord_flip() +
  scale_fill_brewer(palette = "Blues") +
  labs(
    title = "Top 10 Most Competitive States (by Median Raw Margin)",
    x = "State",
    y = "Absolute Vote Margin"
  ) +
  theme_minimal() +
  theme(legend.position = "none")

```

<br> By raw margins, the states of Nevada, Delaware, New Hampshire seem
to be consistent in how much their states are consistently close when it
comes to Presidential elections. This is raw votes. <br>

**The benefits of targeting these kind of states are:**

*1)* Being able to flip the state while targeting fewer voters.

*2)* Because these states are smaller, they are generally less crowded
with national media attention than the bigger states.

**Some of the negatives of targeting these states include:**

*1)* Fewer electoral votes, because these states are smaller in
populations

*2)* Volatility in that small states can swing fast due to miniscule
demographic shifts, third-party influence, or localized issues

*3)* Targeting these states as a main strategy lowers your ROI, because
a marginal win in these smaller states is done at the expense of losing
a much larger and influential swing state with more electoral votes.

<br> ***Most Competitive States by Percentage Margins*** <br>

```{r, warning = FALSE, echo = FALSE}
state_margins <- decadesdata %>%
  filter(party_simplified %in% c("DEMOCRAT", "REPUBLICAN")) %>%
  group_by(year, state, party_simplified) %>%
  summarise(votes = sum(candidatevotes), .groups = "drop") %>%
  pivot_wider(names_from = party_simplified, values_from = votes, values_fill = 0) %>%
  mutate(
    total_votes = DEMOCRAT + REPUBLICAN,
    abs_margin = abs(DEMOCRAT - REPUBLICAN),
    pct_margin = abs_margin / total_votes * 100,
    winner = if_else(REPUBLICAN > DEMOCRAT, "REPUBLICAN", "DEMOCRAT")
  ) %>%
  select(year, state, winner, abs_margin, pct_margin)

top10_abs <- state_margins %>%
  group_by(state) %>%
  summarise(mean_abs_margin = mean(abs_margin, na.rm = TRUE)) %>%
  slice_min(mean_abs_margin, n = 10) %>%
  pull(state)

top10_abs_summary <- state_margins %>%
  group_by(state) %>%
  summarise(mean_abs_margin = mean(abs_margin, na.rm = TRUE)) %>%
  slice_min(mean_abs_margin, n = 10)

top10_median_pct <- state_margins %>%
  group_by(state) %>%
  summarise(median_pct_margin = median(pct_margin, na.rm = TRUE)) %>%
  arrange(desc(median_pct_margin)) %>%
  slice(1:10)

# Plot
state_margins %>%
  filter(state %in% top10_median_pct$state) %>%
  mutate(state = factor(state, levels = top10_median_pct$state)) %>%
  ggplot(aes(x = state, y = pct_margin, fill = state)) +
  geom_boxplot() +
  coord_flip() +
  scale_fill_brewer(palette = "Reds") +
  labs(
    title = "Top 10 Most Competitive States (by Median % Margin)",
    x = "State",
    y = "Margin (%)"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```

<br> Nevada yet again, comes out on top with being the most competitive
state, this time by percentage margin. All of these states have
electoral college votes of 7 or more, and 5 of the 7 modern day swing
states are on this top 10 list, showing that since at least 1976, they
have consistently been among the most competitive states.

**Some of the benefits of targeting these kinds of states in
Presidential Elections include:**

*1)* Scalable impact because even small changes in overall persuasion
can shift the outcome.

*2)* Cost-effective, because of a higher ROI. More money spent but with
the benefit of a higher electoral vote count.

*3)* Electoral College Weight, because these states have more electoral
votes than smaller states.

**Some of the negatives of targeting these states would be:**

*1)* More raw vote totals needed, so even if you might get a higher
electoral vote count, you do it at the expense of more money invested.

*2)* The nature of these states being this close and because of their
total electoral counts means more attention from both parties, which
means voters become inundated with nonstop media coverage, ads, and
canvassers, causing potential voter fatigue.

### Least Competitive States

For this section, we are just going to look at the inverse of the most
competitive states.

```{r, warning = FALSE, echo = FALSE}
state_margins <- decadesdata %>%
  filter(party_simplified %in% c("DEMOCRAT", "REPUBLICAN")) %>%
  group_by(year, state, party_simplified) %>%
  summarise(votes = sum(candidatevotes), .groups = "drop") %>%
  pivot_wider(names_from = party_simplified, values_from = votes, values_fill = 0) %>%
  mutate(
    total_votes = DEMOCRAT + REPUBLICAN,
    abs_margin = abs(DEMOCRAT - REPUBLICAN),
    pct_margin = abs_margin / total_votes * 100,
    winner = if_else(REPUBLICAN > DEMOCRAT, "REPUBLICAN", "DEMOCRAT")
  ) %>%
  select(year, state, winner, abs_margin, pct_margin)

top10_abs <- state_margins %>%
  group_by(state) %>%
  summarise(mean_abs_margin = mean(abs_margin, na.rm = TRUE)) %>%
  slice_min(mean_abs_margin, n = 10) %>%
  pull(state)

top10_abs_summary <- state_margins %>%
  group_by(state) %>%
  summarise(mean_abs_margin = mean(abs_margin, na.rm = TRUE)) %>%
  slice_min(mean_abs_margin, n = 10)

bottom10_median_abs <- state_margins %>%
  group_by(state) %>%
  summarise(median_abs_margin = median(abs_margin, na.rm = TRUE)) %>%
  arrange(desc(median_abs_margin)) %>%
  slice(1:10)

state_margins %>%
  filter(state %in% bottom10_median_abs$state) %>%
  mutate(state = factor(state, levels = rev(bottom10_median_abs$state))
         )%>%
  ggplot(aes(x = state, y = abs_margin, fill = state)) +
  geom_boxplot() +
  coord_flip() +
  scale_fill_brewer(palette = "Blues") +
  scale_y_continuous(labels = scales::comma) +
  labs(
    title = "Top 10 Least Competitive States (by Median Raw Margin)",
    x = "State",
    y = "Absolute Vote Margin"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```

<br> By raw votes, New York, California, and Texas are the least
competitive states. Not surprising, because what do they have in common?
The top 3 least competitive states are in the top 4 of population in the
United States. This is why looking at the margins by percentage is just
as, if not more important, because context matters.

```{r, warning = FALSE, echo = FALSE}
state_margins <- decadesdata %>%
  filter(party_simplified %in% c("DEMOCRAT", "REPUBLICAN")) %>%
  group_by(year, state, party_simplified) %>%
  summarise(votes = sum(candidatevotes), .groups = "drop") %>%
  pivot_wider(names_from = party_simplified, values_from = votes, values_fill = 0) %>%
  mutate(
    total_votes = DEMOCRAT + REPUBLICAN,
    abs_margin = abs(DEMOCRAT - REPUBLICAN),
    pct_margin = abs_margin / total_votes * 100,
    winner = if_else(REPUBLICAN > DEMOCRAT, "REPUBLICAN", "DEMOCRAT")
  ) %>%
  select(year, state, winner, abs_margin, pct_margin)

bottom10_median_pct <- state_margins %>%
  group_by(state) %>%
  summarise(median_abs_margin_pct = median(pct_margin), na.rm = TRUE) %>%
  arrange(desc(median_abs_margin_pct)) %>%
  slice(1:10)

state_margins %>%
  filter(state %in% bottom10_median_pct$state) %>%
  mutate(state = factor(state, levels = rev(bottom10_median_pct$state))) %>%
  ggplot(aes(x = state, y = pct_margin, fill = state)) +
  geom_boxplot() +
  coord_flip() +
  scale_fill_brewer(palette = "Reds") +
  scale_y_continuous(labels = function(x) paste0(x, "%")) +
  labs(
    title = "Bottom 10 Least Competitive States (by Median Percentage Margin)",
    x = "State",
    y = "Vote Margin (%)"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```

<br> Not a surprise, the District of Columbia is the least competitive
set of electoral votes in the country, at least since 1976. Wyoming,
Utah, Idaho, and Oklahoma also seem to be states that are not typically
competitive and are reliable to vote in a single direction.
